\pkg{ggenealogy}: An \proglang{R} Package for Visualizing Genealogical Data

### 2016-11-22 article.tex
Formatting

%% an abstract and keywords
\Abstract{
  This paper introduces \pkg{ggenealogy} (\citealt{ggen}), a developing \proglang{R} software package that provides tools for searching through genealogical data, generating basic statistics on their graphical structures using parent and child connections, and displaying the results. It is possible to draw the genealogy in relation to variables related to the nodes, and to determine and display the shortest path distances between the nodes. Production of pairwise distance matrices and genealogical diagrams constrained on generation are also available in the visualization toolkit. The tools are being tested on a dataset with milestone cultivars of soybean varieties (\citealt{soybean}) as well as on a web-based database of the academic genealogy of mathematicians (\citealt{mgp}). The latest stable package version is available in source and binary form on the Comprehensive \proglang{R} Archive Network.}

@@

Most of these plotting tools were developed by building upon the \pkg{ggplot2} package (\citealt{ggplot2}). As such, most of these plotting tools can be customized by appending syntax from the \pkg{ggplot2} package, as will be demonstrated in demonstrative code throughout this paper. Moreover, some of these plotting tools have interactive capabilities that were developed with the use of the \pkg{plotly} package (\citealt{plotly}) (see Subsection \ref{interaction}).

@@

However, some \pkg{ggenealogy} plots also make use of quantitative variable values associated with individuals in the genealogy. For these plots, the input data frame should also contain a third quantitative variable column. Each case in this quantitative variable column should be of type numeric. In the first example dataset, columns that could be used for this purpose (once subsetted to filter out values of ``'' and NA) include ``devYear", ``yield", and ``yearImputed"; in the second example dataset, the column that could be used for this purpose is ``gradYear". At this point, any \pkg{ggenealogy} plot can be used.

@@

We would like to emphasize that the \code{plotPathOnAll()} method can be created with one or two quantitative variables that the user hard-codes. We demonstrate this using the filtered data frame and igraph objects (\code{sbFilt} and \code{sbFiltIG}) that we previously created by subsetting the soy bean genealogy data to only include observations that contain a value for the quantitative variable ``yield". Then, we can examine the path across the entire genealogical structure and how the quantitative variable ``yield" changes along the parent-child relationships of that path. Likewise, we can plot how the quantitative variables ``devYear" and ``yield" both change along the parent-child relationships of that path. The output of these two calls to the \code{plotPathOnAll()} function can be viewed in Figure \ref{fig:plotPathOnAllQV2}.

@@

\begin{Code}
>R varieties <- c("Brim", "Bedford", "Calland", "Dillon", "Hood", "Narow",
+    "Pella", "Tokyo", "Young", "Zane")
>R plotDegMatrix(varieties, sbIG, sbGeneal) +
+    ggplot2::scale_fill_continuous(low = "white", high = "darkgreen") +
+    ggplot2::theme(legend.title = ggplot2::element_text(size = 15),
+    legend.text = ggplot2::element_text(size = 15)) + ggplot2::labs(x = 
+    "Variety", y = "Variety")
\end{Code}

@@

We see from Figure \ref{fig:dCox} that Sir \code{David Cox} had 42 ``children", many of them becoming notable statisticians themselves, such as Basilio Pereira, Valerie Isham, Gauss Cordeiro, Peter McCullagh, and Henry Wynn. Of his ``children", the one who produced the most ``children" of their own was Peter Bloomfield, who has 26 ``children" and 49 ``descendants". In total, Sir \code{David Cox} had five generations of academic statistics mentees in this dataset.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> statIG <- dfToIG(statGeneal)
R> pathCB <- getPath("David Cox", "Petra Buzkova", statIG, statGeneal,
+    "gradYear", isDirected = FALSE)
R> plotPath(pathCB, statGeneal, "gradYear", fontFace = 4) +
+    ggplot2::xlab("Graduation Year") + ggplot2::theme(axis.text = 
+    ggplot2::element_text(size = 10), axis.title = ggplot2::element_text(
+    (size = 10)) + ggplot2::scale_x_continuous(expand = c(.1, .2))
\end{CodeInput}
\end{CodeChunk}

@@

We could still improve upon Figure \ref{fig:plotCBNoText}. Even though we may be primarily interested in understanding how the path of interest is overlaid across the entire genealogical structure, we could, upon viewing the entire structure, also develop an interest in nodes that are not on the path of interest but are revealed to stand out among the rest of the genealogical structure. For instance, in Figure \ref{fig:plotCBNoText}, it may be of interest for us to determine the names of the few individuals who obtained their statistics degrees before 1900. Fortunately, within the \code{plotPathOnAll()} function, there is a variable \code{animate} that we can set to a value of TRUE to create an interactive version of the figure that allows us to hover over individual illegible labels and immediately receive their labels in a readable format. This interactive functionality comes from methods in the \pkg{plotly} package (\citealt{plotly}). A short video demonstration of these interactive features can be viewed below.

@@

Answering these questions in a straightforward manner requires more than basic data frame manipulation: It also requires methods that can easily traverse parent-child relationships. The \pkg{ggenealogy} package has two methods that can answer these questions using branch traversal. The\code{getBranchQuant()} function can be used to track a quantitative variable across branches and the \code{getBranchQual()} method can be used to track a qualitative variable across branches.

\subsection{Quantitative variable parsing and calculations}

We can demonstrate the \code{getBranchQuant()} function by examining the quantitative variable ``yield" across the descendant branches of the soybean variety \code{A.K.}. To understand more about the output of this function, please consult the \pkg{ggenealogy} package documentation. In the code below, we remove the output column ``DesNames" because it verbosely lists all descendant names, which is not necessary for this demonstration.

@@

We see from the output that \code{A.K.} has two children named \code{A.K. (Harrow)} and \code{Illini}. Descendants from the \code{A.K. (Harrow)} branch have a higher mean yield than the \code{Illini} branch (2932.154 versus 2856.667). However, we should recognize that even though the branches contain a large number of descendants (54 and 131), most of these descendants did not come with a yield value (41 and 125). As a result, the mean values are calculated from a small proportion of the descendants.

@@

\subsection{Qualitative variable parsing and calculations}

The \code{getBranchQual()} function requires similar inputs as the \code{getBranchQuant()} function, except that it also requires an input parameter called \code{rExpr}. The user must initialize this input parameter to a regular expression that can be applied to the column containing the qualitative variable of interest. The regular expression syntax must work on a data frame column of type character. It must be saved as a double quotation string, and any quotation marks within it must be single quotations. The term ``geneal\$colName" must be used in the regular expression.

@@

We will demonstrate one more example of the \code{getBranchQual()} function by searching the qualitative variable ``school" across the ``descendant" branches of the academic statistician \code{David Cox}. The Mathematics Genealogy Project coding system for the ``school" variable was non-ambiguous, and so we do not have to worry about all the various ways the same school could be coded in the dataset. As a result, we now do not have to search for various substrings; we can simply use a regular expression that equates to one value.

It may be interesting to examine the school that is represented the most among all descendants of \code{David Cox}. To determine what school this is, we use the \code{getDescendants()} function to create a data frame called \code{desDC} that contains the names of all 159 ``descendants" of \code{David Cox}. Then, we use the base R function \code{match()} to match the school names from the original dataset to each of the 159 ``descendants" in the \code{desDC} data frame. After that, we use the base R functions \code{sort()} and \code{table()} to examine the five schools that were represented the most throughout the 159 ``descendants".

@@

The examples above demonstrate that users can quickly and flexibly parse descendant branches. The swiftness comes from \pkg{ggenealogy} functions that allow for fast parent-child traversals, such as \code{getChild()}, \code{getDescendants()}, \code{getBranchQuant()}, and \code{getBranchQual()}. The flexibility comes from data frame manipulation functions in base R that can be used in conjunction with the parent-child traversals.

@@

\section*{Acknowledgments}

>>>>

%% an abstract and keywords
\Abstract{
  This paper introduces \pkg{ggenealogy} (\citealt{ggen}), a developing \proglang{R} software package that provides tools for searching through genealogical data, generating basic statistics on their graphical structures using parent and child connections, parsing and performing calculations on branches of interest, and displaying the results. It is possible to draw the genealogy in relation to variables related to the nodes, and to determine and display the shortest path distances between the nodes. Production of pairwise distance matrices and genealogical diagrams constrained on generation are also available in the visualization toolkit. The tools are being tested on a dataset with milestone cultivars of soybean varieties (\citealt{soybean}) as well as on a web-based database of the academic genealogy of mathematicians (\citealt{mgp}). The latest stable package version is available in source and binary form on the Comprehensive \proglang{R} Archive Network.}

@@

Most of these plotting tools were developed by building upon the \pkg{ggplot2} package (\citealt{ggplot2}). As such, most of these plotting tools can be customized by appending syntax from the \pkg{ggplot2} package, as will be demonstrated in example code throughout this paper. Moreover, some of these plotting tools have interactive capabilities that were developed with the use of the \pkg{plotly} package (\citealt{plotly}) (see Subsection \ref{interaction}).

@@

However, some \pkg{ggenealogy} plots also make use of quantitative variable values associated with individuals in the genealogy. For these plots, the input data frame should also contain a third quantitative variable column. Each case in this quantitative variable column should be of type numeric. In the first example dataset, columns that could be used for this purpose include ``devYear", ``yield", and ``yearImputed"; in the second example dataset, the column that could be used for this purpose is ``gradYear". However, for these quantitative variable columns to successfully be used in plots, we would first need to assure that each row within them contains a numeric value (not NA). We could achieve this by filtering out or imputing certain rows that contain NA values for this column of interest. We demonstrate a filtering process for this purpose at the end of Section \ref{psp}. After that, any \pkg{ggenealogy} plot can be used.

@@

We would like to emphasize that the \code{plotPathOnAll()} method can be created with one or two quantitative variables that the user hard-codes. We demonstrate this using the filtered data frame and igraph objects (\code{sbFilt} and \code{sbFiltIG}) that we previously created to assure that there were no NA values in the quantitative variable ``yield". Then, we can examine the path across the remaining genealogical structure and how the quantitative variable ``yield" changes along the parent-child relationships of that path. Likewise, we can plot how the quantitative variables ``devYear" and ``yield" both change along the parent-child relationships of that path. The output of these two calls to the \code{plotPathOnAll()} function can be viewed in Figure \ref{fig:plotPathOnAllQV2}.

@@

\begin{Code}
>R varieties <- c("Brim", "Bedford", "Calland", "Dillon", "Hood", "Narow",
+    "Pella", "Tokyo", "Young", "Zane")
>R plotDegMatrix(varieties, sbIG, sbGeneal) + ggplot2::scale_fill_continuous
+    (low = "white", high = "darkgreen") + ggplot2::theme(legend.title =
+    ggplot2::element_text(size = 15), legend.text = ggplot2::element_text
+    (size = 15)) + ggplot2::labs(x = "Variety", y = "Variety")
\end{Code}

@@

We see from Figure \ref{fig:dCox} that Sir \code{David Cox} had 42 ``children", many of them becoming notable statisticians themselves, such as \code{Basilio Pereira}, \code{Valerie Isham}, \code{Gauss Cordeiro}, \code{Peter McCullagh}, and \code{Henry Wynn}. Of his ``children", the one who produced the most ``children" of their own was \code{Peter Bloomfield}, who has 26 ``children" and 49 ``descendants". In total, Sir \code{David Cox} had five generations of academic statistics mentees in this dataset.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> statIG <- dfToIG(statGeneal)
R> pathCB <- getPath("David Cox", "Petra Buzkova", statIG, statGeneal,
+    "gradYear", isDirected = FALSE)
R> plotPath(pathCB, statGeneal, "gradYear", fontFace = 4) + ggplot2::xlab
+    ("Graduation Year") + ggplot2::theme(axis.text = ggplot2::element_text
+    (size = 10), axis.title = ggplot2::element_text(size = 10)) +
+    ggplot2::scale_x_continuous(expand = c(.1, .2))
\end{CodeInput}
\end{CodeChunk}

@@

We could still improve upon Figure \ref{fig:plotCBNoText}. Even though we may be primarily interested in understanding how the path of interest is overlaid across the entire genealogical structure, we could, upon viewing the entire structure, also develop an interest in nodes that are not on the path of interest but are revealed to stand out among the rest of the genealogical structure. For instance, in Figure \ref{fig:plotCBNoText}, it may be of interest for us to determine the names of the few individuals who obtained their statistics degrees before 1900. Fortunately, within the \code{plotPathOnAll()} function, there is a variable \code{animate} that we can set to a value of TRUE to create an interactive version of the figure that allows us to hover over individual illegible labels and immediately receive their labels in a readable format. This interactive functionality comes from methods in the \pkg{plotly} package (\citealt{plotly}). A short video demonstration of these interactive features can be viewed in Figure \ref{fig:plotAnimate}.

@@

Answering these questions in a straightforward manner requires more than basic data frame manipulation: It also requires methods that can easily traverse parent-child relationships. The \pkg{ggenealogy} package has two methods that can answer these questions using branch traversal. The \code{getBranchQuant()} function can be used to track a quantitative variable across branches and the \code{getBranchQual()} method can be used to track a qualitative variable across branches.

\clearpage
\subsection{Quantitative variable parsing and calculations}

We can demonstrate the \code{getBranchQuant()} function by examining the quantitative variable ``yield" across the descendant branches of the soybean variety \code{A.K.} To understand more about the output of this function, please consult the \pkg{ggenealogy} package documentation. In the code below, we remove the output column ``DesNames" because it verbosely lists all descendant names, which is not necessary for this demonstration.

@@

We see from the output that \code{A.K.} has two children named \code{A.K. (Harrow)} and \code{Illini}. Descendants from the \code{A.K. (Harrow)} branch have a higher mean yield than the \code{Illini} branch (2932.154 versus 2856.667). However, we should recognize that even though the branches contain a large number of descendants (54 and 131), most of these descendants did not come with a yield value (41 and 125). As a result, the mean values were calculated from a small proportion of the descendants.

@@

\clearpage
\subsection{Qualitative variable parsing and calculations}

The \code{getBranchQual()} function requires similar inputs as the \code{getBranchQuant()} function above, except that it also requires an input parameter called \code{rExpr}. The user must initialize this input parameter to a regular expression that can be applied to the column containing the qualitative variable of interest. The regular expression syntax must work on a data frame column of type character. It must be saved as a double quotation string, and any quotation marks within it must be single quotations. The term \code{geneal\$colName} must be used in the regular expression.

@@

We will demonstrate one more example of the \code{getBranchQual()} function by searching the qualitative variable ``school" across the ``descendant" branches of the academic statistician \code{David Cox}. The Mathematics Genealogy Project coding system for the ``school" variable was non-ambiguous, and so we do not have to worry about all the various ways the same school could be coded in the dataset. As a result, we no longer have to search for various substrings; we can simply use a regular expression that equates to one value.

It may be interesting to examine the school that is represented the most among all descendants of \code{David Cox}. To determine what school this is, we use the \code{getDescendants()} function to create a data frame called \code{desDC} that contains the names of all 159 ``descendants" of \code{David Cox}. Then, we use the base R function \code{match()} to match the school names from the original genealogy dataset to each of the 159 ``descendants" in the \code{desDC} data frame. After that, we use the base R functions \code{sort()} and \code{table()} to examine the five schools that were represented the most throughout the 159 ``descendants".

@@

The examples above demonstrate that users can quickly and flexibly parse descendant branches. The swiftness comes from \pkg{ggenealogy} functions that allow for fast parent-child traversals, such as \code{getChild()}, \code{getDescendants()}, \code{getBranchQuant()}, and \code{getBranchQual()}. The flexibility comes from data frame manipulation functions in base R that can be used in conjunction with the parent-child traversal methods.

@@

\clearpage
\section*{Acknowledgments}
